---
title: "A2 Part 3 - Cardio Risk Factors"
format: html
author: Ihina, Nikhita, Rishi
---

# Dataset: Cardio Risk Factors

## Data Dictionary

|  |  |  |
|------------------------|------------------------|------------------------|
| **Variable** | **Description (Cardiovascular Risk Context)** | **Typical Units / Notes** |
| age | Age of the individual; risk generally increases with age | years |
| bmi | Body Mass Index; higher BMI is a risk factor for heart disease | kg/m² |
| waisthip | Waist-to-hip ratio; central obesity increases cardiovascular risk | ratio (waist ÷ hip) |
| smok | Smoking status; major modifiable risk factor | 0 = non-smoker, 1 = smoker (or categorical) |
| choles | Total cholesterol; elevated levels linked to atherosclerosis | mg/dL or mmol/L |
| trig | Triglycerides; high levels can contribute to heart disease | mg/dL or mmol/L |
| hdl | High-density lipoprotein (“good cholesterol”); protective against heart disease | mg/dL or mmol/L |
| ldl | Low-density lipoprotein (“bad cholesterol”); high levels increase risk | mg/dL or mmol/L |
| sys | Systolic blood pressure; high values = hypertension risk | mmHg |
| dia | Diastolic blood pressure; high values = hypertension risk | mmHg |
| uric | Uric acid level; elevated levels can be associated with cardiovascular disease | mg/dL or µmol/L |
| sex | Biological sex; men and women have different risk profiles | 0 = female, 1 = male |
| alco | Alcohol consumption; excessive intake increases risk | units/week or categorical (yes/no) |
| apoa | Apolipoprotein A; component of HDL, protective against heart disease | mg/dL |

In most cardiovascular disease datasets sex is a binary variable coded as 0 and 1, usually, 0 is the reference category and 1 represents the comparison group.

Women are frequently used as the baseline in cardiovascular research because males tend to show higher baseline prevalence and earlier onset of CVD.

Thus we have assumed 0= Female, 1 = Male

### Setting up the Libraries:

```{r}
library(tidyverse)
library(mosaic) # Our trusted friend
library(skimr)
library(ggformula) # Formula interface to ggplot2
library(vcd) # Michael Friendly's package, Visualizing Categorical
library(vcdExtra) # Categorical Data Sets
library(resampledata) # More datasets
library(visStatistics) # Comprehensive all-in-one stats viz/test package
library(ggmosaic) # Tidy Mosaic Plots
library(ggpubr) # Colours, Themes and new geometries in ggplot
library(janitor) # Data Cleaning and Tidying
library(visdat) # Visualize whole dataframes for missing data
library(naniar) # Clean missing data
library(DT) # Interactive Tables for our data
library(tinytable) # Elegant Tables for our data
library(ggrepel) # Repelling Text Labels in ggplot
library(marquee) # Marquee Text in ggplot
library(knitr)
```

### Reading the Data:

```{r}
cardioriskfactors <- CardioDataSets::cardioRiskFactors_df
cardioriskfactors
```

```{r}
glimpse(cardioriskfactors)
```

### Checking for missing data:

```{r}
vis_dat(cardioriskfactors)
vis_miss(cardioriskfactors)
```

### Mutating the data:

```{r}
cardio_mod1 <- cardioriskfactors %>%
  mutate(across(where(is.character), as.factor)
  ) %>% 
  mutate(sex = as.factor(sex),
    smok=as.factor(smok))
cardio_mod1
```

### Correlation between Age and Cholesterol:

```{r}
cardio_mod1 %>%
gf_point(choles ~ age, data = cuisines_modified, color = 'maroon', size = 2) %>%
  gf_lm(color = 'black') %>% 
  gf_labs(title = "Correlation Between Age and Cholesterol",
          x = "Age",
          y = "Total Cholesterol")

```

There is a slight positive correlation between age and cholesterol. More the age, higher the cholesterol. The data also seems to be spread out, great variability.

### Age Distribution:

```{r}
ggplot(cardio_mod1, aes(x = age)) +
  geom_histogram(fill = "#69b3a2", color = "black", bins = 32) +
  labs(
    title = "Age Distribution of people",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

Count of people across ages is fairly distributed.

### Association between sex and smoking status:

```{r}
vcd::structable(data = cardio_mod1, sex ~ smok) %>% 
  as.matrix() %>%  
  addmargins()
```

```{r}
vcd::structable(sex ~ smok, data = cardio_mod1) %>%
    vcd::mosaic(gp = shading_max)
```

There is a significant association between sex and smoking status.

This graph suggests that there are far more non-smoking females than expected as well as more smoking males than expected.

Males are significantly more likely to be smokers than females in this dataset

### BMI Distribution by Sex:

```{r}
ggplot(cardio_mod1, aes(x = factor(sex), y = bmi, fill = factor(sex))) +
  geom_boxplot(alpha = 0.8) +
  labs(
    title = "BMI Distribution by Sex",
    x = "Sex (0 = Female, 1 = Male)",
    y = "Body Mass Index (BMI)"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

```

According to this graph, females have higher BMI than males, and a more varied distribution + more outliers.

### Smoking Status and Sex:

```{r}
cardio_mod1 %>%
  gf_boxplot(sys ~ smok | sex, orientation = "x", fill = ~factor(sex)) %>%
  gf_labs(
    title = "Blood Pressure by Smoking Status and Sex",
    x = "Smoking Status",
    y = "Systolic BP",
    fill = "Sex: 0 = Female, 1 = Male"
  ) %>% 
  gf_refine(scale_fill_brewer(palette = "Pastel1", direction= -1))

```

**Median Blood Pressure Levels:**

-   Female Non-smokers have a slightly higher median (around 130 mmHg) compared to smokers (around 135 mmHg).

-   Males, both non-smokers and smokers have the same median blood pressure (\~137-138 mmHg)

**Sex difference:**

-   Males have higher systolic blood pressure than females across both smoking categories.

-   The distribution patterns are more similar between male groups than between female groups.

The lack of strong smoking effect is surprising, as smoking is typically associated with cardiovascular risks.

```{r}
cardio_mod1 %>% 
  gf_boxplot(choles ~ factor(smok) | factor(sex),
             fill = ~factor(smok),
             orientation = "x") %>%  
  gf_labs(
    title = "Cholesterol by Smoking Status and Sex",
    x = "Smoking Status",
    y = "Cholesterol Level",
    fill = "Smoking Status"
  )

```

**Median Cholesterol Levels:**

-   For females (0): Non-smokers and smokers have almost the same median cholesterol levels (around 5.7)

-   For males (1): Non-smokers have a slightly higher median (around 5.5) compared to smokers (around 5.2

    **Distribution:**

-   Female non-smokers show greater variability in cholesterol levels with 3 outliers above 10 mg/DL

-   Male non-smokers also display more spread in their data compared to male smokers. Smokers (both sexes) have fewer extreme values

    **Sex Differences:** Females generally have higher cholesterol levels than males across both smoking categories. Females show more cases of upper outliers. The shorter line (whisker) for female smokers indicates a smaller range of cholesterol values compared to female non-smokers.

## Summary:

-   Males demonstrate substantially higher smoking rates than females, while females exhibit elevated BMI, more variable cholesterol distributions, and greater outlier prevalence across multiple health metrics. 

-   Age shows a positive correlation with cholesterol levels. 

-   Males have higher systolic blood pressure overall, though females display more blood pressure outliers. 

-   Surprisingly, smoking status shows minimal impact on blood pressure and cholesterol levels within sex groups, suggesting other factors may play stronger roles in cardiovascular health outcomes. 

-   The age distribution remains fairly uniform across the sample population.
